---
title: "Covid19 Time Series Data Analysis"
author: "Alana Hodge"
date: "2023-05-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Import Data

```{r libraryPackages, echo=TRUE, message = FALSE}
library(tidyverse)
library(lubridate)
```

Importing John Hopkins COVID-19 time series Data:

* Global COVID-19 Cases
* Global COVID-19 Deaths
* US-specific COVID-19 Cases
* US-specific COVID-19 Deaths


```{r ImportCovid19Data}
global_cases = read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")
global_deaths = read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

us_cases = read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")
us_deaths = read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")

```

## 2. Tidy And Transform The Data

### Remove unnecessary columns

We most likely don't need the Latitude and Longitude Coordinates in any of the datasets.
We can probably get rid of "iso2", "iso3", "code2" and "FIPS" columns in the US datasets as well.

```{r removeUnnecessaryColumns}
global_cases = global_cases <- subset (global_cases, select = -Long)
global_cases = global_cases <- subset (global_cases, select = -Lat)

global_deaths = global_deaths <- subset (global_deaths, select = -Long)
global_deaths = global_deaths <- subset (global_deaths, select = -Lat)

us_cases = us_cases <- subset (us_cases, select = -iso2)
us_cases = us_cases <- subset (us_cases, select = -iso3)
us_cases = us_cases <- subset (us_cases, select = -code3)
us_cases = us_cases <- subset (us_cases, select = -FIPS)
us_cases = us_cases <- subset (us_cases, select = -Lat)
us_cases = us_cases <- subset (us_cases, select = -Long_)


us_deaths = us_deaths <- subset (us_deaths, select = -iso2)
us_deaths = us_deaths <- subset (us_deaths, select = -iso3)
us_deaths = us_deaths <- subset (us_deaths, select = -code3)
us_deaths = us_deaths <- subset (us_deaths, select = -FIPS)
us_deaths = us_deaths <- subset (us_deaths, select = -Lat)
us_deaths = us_deaths <- subset (us_deaths, select = -Long_)
```

### Handle Missing Data
Next, we handle missing data, by removing any entries that have missing values
in the remaining columns.
```{r handleMissingData}
us_cases <- na.omit(us_cases)
us_deaths <- na.omit(us_deaths)
global_deaths <- na.omit(global_deaths)
global_cases <- na.omit(global_cases)
```


## 3. Visualizations and Analysis

Question: How many people died from COVID-19 during the pandemic years in
North American countries?

```{r NorthAmericanDeathsVisualization, include=TRUE}

q1_data = global_deaths %>% filter(`Country/Region` == "Canada" |
                                     `Country/Region` == "US" |
                                     `Country/Region` == "Mexico")

q1_df1 = q1_data

q1_df1 <- subset (q1_data, select = -`Province/State`)
q1_df1 <- subset (q1_df1, select = -`Country/Region`)
q1_df1$Total_Deaths = rowSums(q1_df1)

q1_data$Total_Deaths = q1_df1$Total_Deaths

q1_plot <- ggplot(q1_data, aes(x = `Country/Region`, y = Total_Deaths)) +
  geom_bar(stat='identity') +
  labs(title = "Number of COVID-19 deaths from 2019-2022 in North America",
       x = "Country",
       y = "Number of deaths")
q1_plot
```


## 4. Modeling



## 5. Bias Implications

Regarding biases identified in the data set:

### Reporting Bias
Not all countries have tested and measured for Covid-19 related cases in deaths in the same way. Specifically for countries with rural populations and/or poor
infrastructure, measuring the actual impact of Covid-19 among the population is
not equitably done across all nations. 

Moreover, it's possible that not all cases of Covid-19 are reported at all.
Especially towards the end of the pandemic, where many people treated their 
symptoms in isolation, it's inaccurate to assume that all cases of Covid-19 in
any given country/region were reported. 

### Personal Bias

The effect of Covid-19 from my own perspective is drastically different from
others. With debates on proper treatment, vaccintaion and policies around
wearing masks, it's reasonable to assume that not everyone reading this report
or utilizing this data has the same perspective on Covid-19 and the pandemic.
This by itself is a form of personal bias that everyone can have towards this
topic, including myself.

## Session info for reproducibility
```{r sessionInfo}
sessionInfo()
```

